---
knit: "bookdown::preview_chapter"
---

# Efficient learning {#learning}

As with any vibrant open source software community, R is fast moving. This can be
disorientating because it means that you can never 'finish' learning R. On the other
hand, it makes R a fascinating subject: there is always more to learn. Even
experienced R users keep finding new functionality that helps solve problems quicker
and more elegantly. Therefore *learning how to learn* is one of the most important
skills to have if you want to learn R *in depth*. We emphasise *depth* of
learning because it is more efficient to learn something properly than to Google it
repeatedly every time you forget how it works.

This chapter aims to equip you with concepts, guidance and tips that will accelerate your transition
from an R *hacker* to an R *programmer*. This inevitably involves effective use of R's
help, reading R source code, and use of online material.

## Top 5 tips for efficient learning

- Use R's internal help, e.g. with `?`, `??`, `vignette()` and `appropos()`. Try **swirl**.
- Read about the latest developments in established outlets such as the *Journal for Statistical Software*, the *R Journal*, R lists and the 'blogosphere'.
- If stuck, ask for help! A clear question posted in an appropriate place, using reproducible code, should get a quick and enlightening answer.
- For more in-depth learning, nothing can beat immersive R books and tutorials. Do some research and decide which resources you should complete.
- One of the best ways to consolidate learning is to write-it-up and pass on the knowledge: telling the story of what you've learned with also help others.

## Using R help

The best place to look for help on R functions is often within R itself. Although R's internal documentation is notoriously terse, studying it as the first 'port of call' when stuck is time well spent: you will see what the function was designed to do and perhaps gain further insight about its history, e.g. through references to academic papers.

All functions exported functions have help files, which contain, at a minimum, a list of the input arguments and the nature of the output that can be expected. Prefixing a function name with `?` will show the help for that function. To the help file for `plot`, for example just type:

```{r eval=FALSE}
# Or help("plot")
?plot
```

In general, help pages describe the public API of the package. This is one of the reasons
that function help pages are thought (by some) to be difficult to understand. In practice, 
this means that the help page does not describe the underlying mathematics or algorithm in 
detail, it's aim is to describe the interface.

A help page is divided into a number of sections. The most _helpful_
section (I find) is the examples (at the bottom of the help page) showing precisely
how the function works. You can either copy and paste the code, or actually run the
example code using the `example` command:.

```{r eval=FALSE}
example(plot)
```

```{block, type="rmdnote"}
When a package is added to CRAN, the example part of the documentation is run on all major
platforms. This helps ensure that a package works on multiple systems.
```

Another useful section in the help file is `See Also:`. In the `plot` help file, it
gives pointers to 3d plotting.

To look for help about a certain *topic* rather than a specific *function* use
`??topic`, which is analogous to `?function`. To search for information about
regression in all installed packages, for example, use the following command:

```{r, eval=FALSE}
# Or help.search("regression")
??regression
```

To search more specifically for objects the `appropos` function can be useful. To
search for all objects and functions in the current workspace containing the text
string `lm`, for example, one would enter:

```{r, echo=-2}
# Showing the first six results
apropos("lm")[1:6]
```

Sometimes a package contains vignettes. To browse any vignettes associated with a
particular package, we can use the handy function

```{r eval=FALSE}
browseVignettes(package = "benchmarkme")
```

### Reading R source code

R is open source. This means that we view the underlying source code and examine any
function. Of course the code is complex, and diving straight into the source code
won't help that much. However, watching to the github R source code
[mirror](https://github.com/wch/r-source/) will allow you to monitor small changes
that occur. This gives a nice entry point into a complex code base. Likewise, examining
the source of small functions, such as `NCOL` is informative, e.g.
`getFunction("NCOL")`

```{block, type="rmdtip"}
Subscribing to the R NEWS [blog](https://developer.r-project.org/blosxom.cgi/R-devel/NEWS/)
is an easy way of keeping track of future changes.
```

Many R packages are developed in the open on github or r-forge. Select a few well
known packages and examine their source. A good package to start with is
**[drat](https://github.com/eddelbuettel/drat)**. This is a relatively simple package
developed by Dirk Eddelbuettel (author of Rcpp) that only contains a few functions. It
gives you an excellent pointer into software development by one of the key R package
writers.

## Online resources

The R community has a strong online presence, providing many resources for learning. The R-project website contains six detailed [official manuals](https://cran.r-project.org/manuals.html), plus a giant pdf file containing documentation for all recommended packages. These include [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html), [The R language definition](https://cran.r-project.org/doc/manuals/r-release/R-lang.html) and [R Installation and Administration](https://cran.r-project.org/doc/manuals/r-release/R-admin.html), all of which are recommended for people wanting to learn their general R skills. If you are developing a package and want to submit it to CRAN, the [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html) manual is recommended reading, although it has to some extent been superseded by @Wickham_2015, the source code of which is [available online](https://github.com/hadley/r-pkgs). While these manuals are long, they contain important information written by experienced R programmers.

The [R-journal](https://journal.r-project.org/) regularly publishes articles
describing new R packages, as well as general programming hints. Similarly, the
articles in the [Journal of Statistical Software](https://www.jstatsoft.org/) have a
strong R bias. Publications in these journals are generally of very high quality and have been rigorously peer reviewed. However, they may be rather technical for R novices.

The wider community provides a much larger body of information, of more variable quality, than the official R resources. The Contributed Documentation page on R's home page contains a dozens of tutorials and other resources on a wide range of topics. Some of these are excellent, although many are not kept up-to-date.

Lower grade but more frequently released information can be found on the 'blogosphere'. Central to this is  [R-bloggers](http://www.r-bloggers.com/),
a blog aggregator of content contributed by bloggers who write about R (in English).
It is a great way to get exposed to new and different packages. Similarly monitoring
the _[#rstats](https://twitter.com/search?q=%23rstats)_ twitter tag keeps you up-to-date
with the latest news.

There are also mailing lists, Google groups and the Stack Exchange Q & A sites. Before
requesting help, read a few other questions to learn the format of the site. Make sure
you search previous questions so you are not duplicating work. Perhaps the most
important point is that people aren't under __any__ obligation to answer your
question. One of the fantastic things about the open-source community is that you can
ask questions and one of core developers may answer your question free; but remember,
everyone is busy!

### Stackoverflow

The number one place on the internet for getting help on programming is
[Stackoverflow](http://www.stackoverflow.com). This website provides a platform for
asking and answering questions. Through site membership, questions and answers are
voted up or down. Users of Stackoverflow earn reputation points when their question or
answer is up-voted. Anyone (with enough reputation) can edit a question or answer.
This helps the content remain relevant.

Questions are tagged. The R questions can be found under the [R tag](http://stackoverflow.com/questions/tagged/r). The [R page](https://stackoverflow.com/tags/r/info) contains links to Official documentation, free resources, and various other links. Members of the Stackoverflow R community have tagged, using `r-faq`, a few question that often crop up. 

### Mailing lists and groups.

There are many mailing lists and Google groups focused on R and particular packages. The main list for getting help is
`R-help`. This is a high volume mailing list, with around a dozen messages per day. A more technical mailing list
is `R-devel`. This list is intended for questions and discussion about code development in R. The discussion on this list is very technical. However,
it's a good place to be introduced to new ideas - but it's not the place to ask about these ideas! There are 
many other special interest mailing [lists](https://www.r-project.org/mail.html) covering topics such as high performance computing to ecology.
Many popular packages also have their own mailing list or Google group, e.g. **ggplot2** and **shiny**. The key piece of advice is before mailing a
list, read the relevant mailing archive and check that your message is appropriate.

### Asking a question

Asking questions on stackoverflow and R-help is hard. Your question should contain
just enough information that you problem is clear and can be reproducible, while
at the same time avoid unnecessary details. Fortunately there is a SO question - [How to make a great R reproducible example?](http://stackoverflow.com/q/5963269/203420) - that provides
excellent guidance!

#### Minimal data set {-}

What is the smallest data set you can construct that will reproduce your issue? Your
actual data set may contain $10^5$ rows and $10^4$ columns, but to get your idea across
you might only need $4$ rows and $3$ columns. Making small example data sets is easy. 
For example, to create a data frame with two numeric columns and a column of characters
just use

```{r echo=-3}
set.seed(1)
example_df = data.frame(x = rnorm(4), y = rnorm(4), z = sample(LETTERS, 4))
```

Note the call to `set.seed` ensures anyone who runs the code will get the same
random number stream. Alternatively, you use one of the many data sets that come with R - `library(help = "datasets")`. 

If creating an example data set isn't possible, then use `dput` on your actual data set. This
will create an ASCII text representation of the object that will enable anyone to recreate
the object

```{r results="hide"}
dput(example_df)
#> structure(list(
#>  x = c(-0.626453810742332, 0.183643324222082, -0.835628612410047, 1.59528080213779), 
#>  y = c(0.329507771815361, -0.820468384118015, 0.487429052428485, 0.738324705129217), 
#>  z = structure(c(3L, 4L, 1L, 2L), .Label = c("J", "R", "S", "Y"), class = "factor")), 
#>  .Names = c("x", "y", "z"), row.names = c(NA, -4L), class = "data.frame")
```

#### Minimal example {-}

What you should not do, is simply copy and paste your entire function into your question.
It's unlikely that your entire function doesn't work, so just simplify it the bare minimum. 
The aim is to target your actual issue. 
Avoid copying and pasting large blocks of code; remove superfluous lines that are not part of the problem.
Before asking your question, can you run you code in a clean R environment and reproduce your error?



